---
title: "Assignment 10"
author: "Taha Ahmad"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r lib import, message=FALSE}
library(tidyverse)
library(pracma)
library(glue)
library(stats)
```

### Problem 1

Smith is in jail and has 1 dollar; he can get out on bail if he has 8 dollars.
A guard agrees to make a series of bets with him. If Smith bets A dollars,
he wins A dollars with probability .4 and loses A dollars with probability .6.

Find the probability that he wins 8 dollars before losing all of his money if:

(a) he bets 1 dollar each time (timid strategy).

(b) he bets, each time, as much as possible but not more than necessary to
bring his fortune up to 8 dollars (bold strategy).

(c) Which strategy gives Smith the better chance of getting out of jail?

#### Information


Taking the timid strategy transforms this into a Gambler's Ruin problem. So we can utilize the Gambler's Ruin formula turned into a function below:

```{r information 1}
p <- 0.4
q <- 1-p
z <- 1
M <- 8
gam_ruin <- function(p,q,z,M) {
  return (
    ((q/p)**z - 1) / ((q/p)**M-1)
  )
}
```

#### Attempt A

If Smith takes the timid strategy he has a **0.0203 probability** of winning his bail before losing all his money.

```{r attempt A}
(gam_ruin(p,q,z,M))
```

#### Attempt B

If Smith takes the bold strategy, he loses all his money after a single failure. This means we are simply looking for the probability of 3 successes in a row without any failures to get to $8. 

If Smith takes the bold strategy he has a **0.064 probability** of winning his bail before losing all his money.

```{r attempt B}
( (p**3)*(q**0) )
```

#### Attempt C

The **bold strategy** gives Smith the better chance of getting out of jail.

