---
title: "Week 3 Assignment"
author: "Taha Ahmad"
date: '`r Sys.Date()`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r package loading, message=FALSE}
library(tidyverse)
```


### Introduction

This week we're focusing on character manipulation and processing with R.

#### Question 1

Using the 173 majors listed in fivethirtyeight.com’s College Majors dataset [https://fivethirtyeight.com/features/the-economic-guide-to-picking-a-college-major/], provide code that identifies the majors that contain either "DATA" or "STATISTICS". 

The csv containing all the majors can be found here: https://raw.githubusercontent.com/fivethirtyeight/data/master/college-majors/majors-list.csv

```{r question 1}
file = r"(https://raw.githubusercontent.com/fivethirtyeight/data/master/college-majors/majors-list.csv)"

majors <- read_csv(file, col_names = TRUE, show_col_types = FALSE)

majors |> 
  filter(str_detect(str_to_upper(Major), "DATA|STATISTICS"))

```

#### Question 2

Write code that transforms the data below:

[1] "bell pepper"  "bilberry"     "blackberry"   "blood orange"
[5] "blueberry"    "cantaloupe"   "chili pepper" "cloudberry"  
[9] "elderberry"   "lime"         "lychee"       "mulberry"    
[13] "olive"        "salal berry"

Into a format like this:

c("bell pepper", "bilberry", "blackberry", "blood orange", "blueberry", "cantaloupe", "chili pepper", "cloudberry", "elderberry", "lime", "lychee", "mulberry", "olive", "salal berry")

We're going to want to transform a string of the first instance of data into a properly formatted array.

```{r question 2}
raw_data <- r"([1] "bell pepper"  "bilberry"     "blackberry"   "blood orange"
[5] "blueberry"    "cantaloupe"   "chili pepper" "cloudberry"  
[9] "elderberry"   "lime"         "lychee"       "mulberry"    
[13] "olive"        "salal berry")"

# Removes the characters that do not help us with further conversion into an array i.e. "[1] ".
processed_data_1 <- str_remove_all(raw_data,r"(\[.+\]\s)")

# Removing the first and last characters as with the current format they will have quotes which we will not split on.
processed_data_2 <- str_sub(processed_data_1,2,-2)

# Splitting the string on the pattern of quotations with white space between them which perfectly matches how the string is structured.
processed_list <- str_split_1(processed_data_2,r"("\s+")")

# Returning the processed list for our viewing pleasure.
processed_list

# Testing that the code transformation is identical in r's eyes.
test_list <- c("bell pepper", "bilberry", "blackberry", "blood orange", "blueberry", "cantaloupe", "chili pepper", "cloudberry", "elderberry", "lime", "lychee", "mulberry", "olive", "salal berry")
identical(processed_list,test_list)
```

#### Question 3

Describe, in words, what these expressions will match:

    - (.)\1\1
    This will match any character that has been repeated three times.
    - "(.)(.)\\2\\1"
    This will match any of the first character in a sequence and any of the second character in a sequence where the same second character comes next and the first character comes again in the fourth place.
    - (..)\1
    This will match a series of any two characters that has been repeated twice.
    - "(.).\\1.\\1"
    This will match any of the first character in a sequence, any character, the same first character, any character, and the same first character in the fifth place.
    - "(.)(.)(.).*\\3\\2\\1"
    This will match any of three characters in a row, with any or none characters in between the three characters with their order reversed.


#### Question 4

Construct regular expressions to match words that:

    - Start and end with the same character.
    ^(.).*\1$
    - Contain a repeated pair of letters (e.g. "church" contains "ch" repeated twice.)
    ^.*(..).*\1.*$
    - Contain one letter repeated in at least three places (e.g. "eleven" contains three "e"s.)
    ^.*(.).*(\1.*){2,}$


### Conclusions

We've mainly focused on regex with this assignment. What I do to enhance my usage of regex is utilize https://regex101.com/ for building regex queries as I can take in a portion of sample data to ensure that the desired data matching is taking place.There is support for multiple different flavors of regex, but not POSIX ERE which R uses by default. PCRE2 from the site will work if you set the regex flavor by calling (perl = true). A breakdown of the differences between regex engines can be found here: https://gist.github.com/CMCDragonkai/6c933f4a7d713ef712145c5eb94a1816.
